---
title: "Analysis of parallel groups designs"
author: "Daniel Hammarström"
date: "17 oktober 2018"
bibliography: lecture07_bib.bib
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse); library(readxl)
#set.seed(1)
#
#df <- data.frame(group = rep(c("A", "B", "C"), each = 10),  outcome = c(rnorm(10, 5, 2), #rnorm(10, 7.8, 2), rnorm(10, 10, 2))) 
#
#avg <- df %>%
#        group_by(group) %>%
#        summarise(m = mean(outcome))
#
#avg.all <- df %>%
#        group_by() %>%
#        summarise(m = mean(outcome))
#
#
#
#df %>%
#        ggplot(aes(group, outcome)) + geom_jitter(width = 0.2) + 
#        geom_point(data = avg, aes(group, m))+
#        geom_segment(data = avg, aes(c(0.75, 1.75, 2.75),
#                                     m, 
#                                     yend = m,
#                                     xend = c(1.25, 2.25, 3.25))) +
#        geom_hline(yintercept = avg.all$m, color = "blue")
#
#
#
#contrasts(df$group) <- "contr.helmert"
#
#
#a <- aov(outcome ~ group, data = df)
#l <- lm(outcome ~ group, data = df)
#?aov
#summary(a)
#summary(l)

```

# Background

- In sport science (and e.g. medical-, nutritional-, psychological-sciences), intervention-studies are common. We are interested in the effect of e.g. a training method, nutritional supplement or drug. 

- The outcome in these studies could be physical performance, degree of symptoms, muscle size or some other measure that we are interested in studying.

- These studies are often called *Randomized Controlled Trials* (RCT)

 
# Different study design

- The choice of study design relates to the research-question and dictates what statistical methods can be applied.

- The study design affects the ability of the study to detect an effect (the power)

- A common case of a RCT is the parallel-groups design

- Participants are **allocated** to two or more "treatment groups", **at random**, one group gets a treatment, the other group acts as the control. 

- Usually, a measurement is made prior to the intervention (Baseline) and after the intervention.

- This is a common design **when wash-out is not possible** and thus, the two treatment can not be compared within the same individual. 

# A pre-post parallel-groups design


```{r, fig.align='center', fig.dim = c(4,2.5)}

data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(0.5,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "Treatment") +
        annotate(geom = "text", x = 2.25, y = 0.5, label = "Control") 

```

# A pre-post parallel-groups design


```{r, fig.align='center', fig.dim = c(4,2.5)}

data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(1.1,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "A") +
        annotate(geom = "text", x = 2.25, y = 1.1, label = "B") 

```


# What is the question

- Here we are interested in the **treatment effect** (or the difference in effect of two different treatments)

- This means that we want to establish if

$$ \Delta Y_A-\Delta Y_B \neq 0 $$

... meaning that the **null hypothesis** is that the **change** ($\Delta$) in group A is not different to the **change** in group B

# Different statistical solutions (1)

- We could do a *t-test* on the change score between groups. This is equivalent to a regression model where we model change as a function of groups

$$outcome = \beta_0 + \beta_1 Group_B$$
\small
```{r, eval = FALSE, echo = TRUE}
# t.test example
with(data, t.test(outcome_A, outcome_B, paired = FALSE)

# The same in regression
lm(change ~ group, data = data)
```
\normalsize

# Problems with the simple solution

- Baseline values can affect the interpretation of a pre- to post-intervention study through **regression to the mean** 

- If we analyse change scores ($post-pre$), regression to the mean will give an overestimation of the effect, if there is, by chance, a difference in baseline values between groups (lower values in treatment group) [@RN2118].

- If we analyse follow up scores, the pattern will be reversed.

# A solution to this problem

- Instead of only analyzing change-scores, we can also control for the relationship between baseline values and the change scores.

- This technique is called Analysis of Co-Variance (ANCOVA), where the baseline is considered the adding the covariance.

- This is an extension of the simple linear regression

\small
```{r, eval = FALSE, echo = TRUE}
# Extending the linear regression equation
lm(change ~ baseline + group, data = data)
```
\normalsize

# Why ANCOVA

- The ANCOVA model has better power [@RN2116]
- The ANCOVA model gives **unbiased** estimates of differences between groups [@RN2118]

# When can we use the ANCOVA model?

- When the allocation of participants have been done at random (e.g. RCTs), differences at baseline should be due to random variation


# Presentation of 10 vs 30RM-study

- 31 participants were assigned to one of two groups training with either 10RM or 30RM, 27 participants completed the trials (24 participants completed to full time).

- The main interest in the study was development of strength and muscle hypertrophy (we are interested in strength)

- The variables in the file are:
a. `subject`: ID of participant
b. `timepoint`: prior to *pre* or after the intervention *post*
c. `group`: The intervention group
d. `exercise`: The exercise that was tested, *legpress*, *benchpress* or *bicepscurl*
e. `load`: The load lifted in 1RM test (kg)

# Reading the data set and exploratory analysis
\footnotesize
```{r, eval = FALSE, echo = TRUE}
library(tidyverse); library()
read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", "post"))) %>%
        ggplot(aes(timepoint, load, fill = group)) + 
        geom_boxplot() +
        facet_wrap(~ exercise, scales = "free")

```
\normalsize

# Exploratory analysis
```{r, fig.align='center', fig.dim = c(4,2.5)}
library(tidyverse); library()
read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        mutate(timepoint = factor(timepoint, levels = c("pre", "post"))) %>%
        ggplot(aes(timepoint, load, fill = group)) + geom_boxplot() +
        facet_wrap(~ exercise, scales = "free") + 
        theme_classic() +
        theme(legend.position="top")
```

# The main purpose of your analysis:

**What training method would you, based on your analysis recommend for improvning maximal strength?**

# How to answer the question

- Choose the most appropriate 1RM test or use all
- Choose the most appropriate statistical model/test, compare different models (`lm` on the change-score, `lm` with baseline as covariate, `lm` on post-values with baseline as a covariate)
- Write a full report (Background, methods (with emphasis on statistics), results and discussion)

# References












