---
title: "Analysis of correlated data - repeated measures designs"
author: "Daniel Hammarström"
date: "23 oktober 2018"
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
output: beamer_presentation
---

# What packages do you need?

```{r, eval = FALSE}
library(readxl) # Read data into R
library(tidyverse) # Data manipulation
library(lme4) # For fitting models
```


# Hypothesis testing

- If we are interested in comparing means between groups or within groups (paired observations), the *t-test* can be used, however
- the t-test is limited to one dependent variable (continuous) and one independent variable (group/condition) and cannot be extended to multiple groups. 
- Multiple testing of more than two groups would violate the specified false discovery rate ($\alpha$)

**We need something else!**


# Repeated measures designs

* Repeated measures designs were typically analyzed using **An**alysis **o**f **Va**riance (ANOVA).
* ANOVA models have a long history and can be used for hypothesis testing comparing means of more than two groups.
* The null-hypothesis of a simple ANOVA is that no group is different to the grand mean ($\mu$):

$$\mu_a = \mu_b = \mu_c = \mu_n = \mu$$

* In the simple case (one-way ANOVA), variation between groups is compared to variations within groups

* In the more complex case (e.g. repeated-measures ANOVA), we can take advantage of variation that is due to the experimental unit (e.g. participant).

* We then partition variance into *known* (e.g. treatment) and *unknown* (residual error) sources.  

# ANOVA, linear regression and limitations

* The ANOVA model is a special case of linear regression, we can therefore test for specific effects after doing the initial test (post-hoc tests) by combining terms from the model
* In the ANOVA, all explaining variables are categorical, we are therefore limited to special study designs
* The typical ANOVA requires balanced data and no missing values.
    - Missing values will require the removal of all data from the specific participant in a repeated measures design
    - Unbalanced data will lead to biased estimates.
    
**We need something else!**

# Extending the linear model - Linear Mixed Models (LMM)

* The LMM is an extension of the linear regression, with (quite a few!) exceptions
* The ordinary linear model assumes independent data
* An ANOVA can account for a single source of independent data, e.g. participants 
* LMM is more flexible than the ANOVA/linear model when dealing with correlated data (e.g. repeated-measures designs).
* Additionally, we can include continuous covariates (e.g. time) analyse more complex designs and we do not have to worry about missing data in the LMM.
* An example of more a complex design could be multiple levels of "unexplained" variation due to e.g. different participants, located at different facilities, coached by different coaches.

# What is a Linear Mixed Model

* The linear mixed model combines to types of **effects**, an effect being a change in the measured quantity due to some event or condition (continuous or categorical).
* A **Fixed effect** is an event/condition were we know all values of interest, this can be a grouping variable that we use in a study design.
* A **Random effect** is an effect that is sampled by random, we do not know all levels of this effect. This can be different participants recruited to a study.
* If we combine **Random** and **Fixed** effects, our model becomes **mixed**

# What is a LMM


* The linear regression model can be written as 

$$y=\beta_0 + \beta_1X_1$$

* We can estimate this model by minimizing the distance from all observations to the "best fit line"
* The mixed model also includes *random effects* that can be thought of as effects specific to e.g. a certain participant

$$y_i=\beta_0 + \beta_1X_1 + b_{i0} \qquad i = 1,...,n$$

* Every participant gets their own intercept!

# Comparing LM to LMM
\scriptsize
```{r, echo = FALSE, warning = FALSE, message=FALSE, results = "asis"}
library(tidyverse); library(readxl); library(nlme); library(lme4); 
library(knitr); library(kableExtra)

dat <- read_excel("./data/ten_vs_thirty_complete.xlsx", na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
        mutate(log.load = log(load)) 



m1a <- lm(load ~ timepoint * group, data = dat)

m1 <- lmer(load ~ timepoint * group + (1|subject), data = dat)


rbind(summary(m1a)$coef[,c(1,2,3)], summary(m1)$coef[,c(1,2,3)]) %>%
        kable(format = "latex", 
              booktabs = TRUE,
              caption = "Regression coefficients from a linear model and a mixed linear model examining the effects of timepoint (within) and group (between) on 1RM strength", digits = 1) %>%
        group_rows("Linear model", 1,6) %>%
        group_rows("Mixed-linear model",7,12)
     


```


\normalsize

# Random intercepts and random slopes

* Random effects are used to account for dependence in the data, i.e. when data is not independent. 
* This will generally improve precision in estimates from the erroneously fitted linear model (we violate assumptions)
* There are two ways we can include random effects, **random slopes** and **random intercepts**
* In a longitudinal design, random intercepts are the deviance for each participant from the over-all intercept, each participant gets their own starting point.
* Random slopes can be added if we think that individuals respond differently to treatments.
* To estimate random effects, we need data to do so. Random intercepts are straight forward, random slopes requires more than two data points to get the estimate.

# Mixed linear models - Examples

* The 10 vs. 30RM-study also had a "mid-point" measurement after 12 weeks.

```{r, echo=FALSE, warning = FALSE, fig.align='center', fig.dim = c(4,2.5)}
library(tidyverse); library(readxl); library(nlme); library(lme4); 
library(knitr); library(kableExtra)

read_excel("./data/ten_vs_thirty_complete.xlsx", na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
        mutate(log.load = log(load)) %>%
        ggplot(aes(timepoint, load, group = subject, color = group)) + geom_line() + 
        theme_classic()

ten_vs_thirty <- read_excel("./data/ten_vs_thirty_complete.xlsx", na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
        mutate(log.load = log(load))

```

# Mixed linear models - Examples

* The data-set contains missing values
* Groups are not balanced
* ANOVA is not an option, we should use mixed models!

# Main effects and interactions

* We want to estimate the effect of time-point, and group. 
* But we are mostly interested in the **interaction** between group and time-point.
* This is the main question in the study, do the groups deviate from each other after the training period?


# How to fit LMM in R
\scriptsize
```{r, eval = FALSE}
library(lme4)

# Fit model
m1 <- lmer(load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty) 

# This is the same as
m1 <- lmer(load ~ timepoint * group + (1|subject),
           data = ten_vs_thirty) 

# Retrieve model coefficients
summary(m1)$coef

```
\normalsize

# Including random effects

* Random effects are included in the model using `(1|subject)`, this means that we have specified a random intercept for each participant.
* A random slope for time could be added to the model if we had more than one measurement at each time-point for each participant or if we treated time as continuous variable.
* A random slope could be added by `(1 + time|subject)`


# How to interpret model coefficients

* Remember the linear regression model, the fixed effects part of an mixed model can be read as an ordinary regression model
* In the example `timepoint` and `group` are categorical variables (coded as 0 and 1)

* Fit the model, how do you read the coefficients?


# Example 
\scriptsize
```{r, eval = FALSE}
ten_vs_thirty <- read_excel("./data/ten_vs_thirty_complete.xlsx", 
                            na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", "mid", "post")))

# Fit model
m1 <- lmer(load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty) 

# Retrieve model coefficients
summary(m1)$coef
```
\normalsize

# What about baseline corrections?

* In the mixed model we just formulated, the `groupRM30` coefficient is difference between groups at the intercept. 
* The difference between groups at the intercept is added to time-point and the effect seen in the interaction term
* Other options for baseline correction could be to model the change scores and include baseline as a covariate

# Interactions

* How do you interpret the interaction term? 

* Make a drawing of the of the model based on the output

```{r, echo = FALSE, warning= FALSE, message = FALSE, results = "asis"}
summary(m1)$coef %>%
        kable(format = "latex", 
              booktabs = TRUE,
              caption = "Regression coefficients", digits = 1)

```




# Model diagnostics

* As mixed linear models are very similar to regression models, we can use some of the same diagnostic tools
* For a first look use `plot(model)`

```{r, message = FALSE, warning = FALSE, echo = FALSE, fig.align='center', fig.dim = c(4,2.5)}

ten_vs_thirty <- read_excel("./data/ten_vs_thirty_complete.xlsx", 
                            na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", "mid", "post"))) %>%
        mutate(log.load = log(load))

# Fit model
m1 <- lmer(load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty) 

plot(m1)


```

# Model diagnostics (2)

```{r, fig.align='center', fig.dim = c(4,2.5)}
qqnorm(resid(m1)); qqline(resid(m1))
```

# Assumptions

* We assume that the residuals from the model are equally scattered over the fitted data (homoscedasticity, check with `plot(model)`)
* We also assume that the residuals are normally distributes (check with `qqnorm(resid(model)); qqline(resid(model))`)

* If we have problems here, transformation of the dependent variable is a first resort

* A efficient transformation is the log transformation


# Mixed models using log-transformed data

\scriptsize

```{r, results = "hide"}
# calculating the log of the dependent variable
ten_vs_thirty <- read_excel("./data/ten_vs_thirty_complete.xlsx", 
                            na = "NA") %>%
        filter(exercise == "legpress") %>%
        spread(timepoint, load) %>%
        gather(timepoint, load, mid:pre) %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", "mid", "post"))) %>%
        mutate(log.load = log(load)) 

# Fitting the model
m1_log <- lmer(log.load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty) 
```

\normalsize

# Model diagnostics

```{r, echo = FALSE, fig.align='center', fig.dim = c(4,4)}
par(mfrow = c(1,2))
qqnorm(resid(m1_log), main = "Log-transformed");qqline(resid(m1_log))
qqnorm(resid(m1), main = "Orginal scale", ylab="");qqline(resid(m1))
```


# Mixed models using log-transformed data

* The interpretation of the log-transformed dependent variable is different to the model with the dependent variable on the original scale.

$$log(A)-log(B) = log(\frac{A}{B})$$

* Back-transformed regression coefficients gives fold-changes, can be interpreted as percentage change.  
* Do not back-transform standard errors, use point-estimates and confidence intervals

\scriptsize
```{r, eval = FALSE}
exp(confint(m1_log))

exp(summary(m1_log)$coef[,1])

```

\normalsize

# Back-transformed estimates

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
cbind(estimate = exp(summary(m1_log)$coef[,1]), exp(confint(m1_log))[c(3:8),]) %>%
        kable(format = "latex", 
              booktabs = TRUE,
              caption = "Regression coefficients", digits = 2)
```



# Inference from mixed models

* Confidence intervals can be calculated on model parameters using `confint(model)`, these can be used for hypothesis testing
* p-values are not provided in `lme4`, this is by design, but there are other ways to calculate p-values for specific effects

# Calculating p-values in the mixed-effects model using likelihood ratio tests

* Likelihood ratio tests (LRT) can be used to calculate p-values. 
* Here we compare a model with an effect, to a model without the effect, the test compares "goodness of fit"

\scriptsize
```{r, eval = FALSE}

# Fit a model with all effects
m1 <- lmer(load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty) 
# Then fit a null model for comparison, we drop the higher order interaction
m0 <- lmer(load ~ timepoint + group + (1|subject),
           data = ten_vs_thirty) 

# compare the models using the anova() function
anova(m0, m1)
```

# LRT p-values

\scriptsize
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}

# Fit a model with all effects
m1 <- lmer(load ~ timepoint + group + timepoint:group + (1|subject),
           data = ten_vs_thirty, REML = FALSE) 
# Then fit a null model for comparison, we drop the higher order interaction
m0 <- lmer(load ~ timepoint + group + (1|subject),
           data = ten_vs_thirty, REML = FALSE) 

# compare the models using the anova() function
anova(m0, m1)
```
\normalsize

# Exercise / Assignment

* We will use the complete 10 vs 30RM data set `ten_vs_thirty_complete.xlsx`
* Compare your analyses of the pre-post data set, with the mixed model approach, do you reach the same conclusions?

