---
title: "Assignments Week 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Possible solutions](assignments_week3_solutions.html)


## Regression models

This week you are expected to perform statistical analysis using regression models. You are expected to perform all assignments below and hand in a .Rmd-file and a .html report (hide the code in the report).

## Problem 1
Using the cycling data set. Fit the following model:

$$VO_{2max} = Height$$

Use data from time-point `meso2`. Check your assumptions and interpret the model. 

## Problem 2

Phosphorylation of the p70 S6 kinase induces increased translational efficiency. In this study (Mitchel et al. 2013), the researcher asked if baseline exercise-induced phosphorylation of p70S6K could explain changes in Cross Sectional Area (CSA) of muscle fibers after a training period.

Methods: Young male participants (n=23) completed a 16 wk training period with 4 weekly strength training sessions. Muscle biopsies were collected from from vastus lateralis before and after the training period. Two biopsies were collected before the training period, before and after an acute exercise session in order to study activation of molecular pathways involved in protein synthesis. Muscle fiber CSA were analyzed from muscle tissue cross sections.  

Data analysis questions: 
1. What is the relationship between phosphorylation status and increase in CSA?
2. How much of the variation in CSA can be explained by p70 S6K?
3. Examine the model fit and check assumptions, identify potential outliers and compare your conclusions to the conslusions reported in the paper.

Reference: Mitchell, C. J., et al. (2013). "Muscular and Systemic Correlates of Resistance Training-Induced Muscle Hypertrophy." PLoS One 8(10): e78636.

### Where to start? 

The data can be found on canvas. After download to your data folder, use the following code to import the data.

```{r, eval = FALSE}
library(tidyverse); library(readxl)

csaP70 <- read_excel("./data/Mitchell2013.xlsx") %>%
        print()
```

## Problem 3

Using the cycling data set. Select a performance variable from the list 

- `VO2.max` (VO~2max~)
- `peak.power` (peak power in wingate test)
- `tte` (time to exhaustion in VO~2max~ test)
- `sj.max` 

Find 2-4 variables that might explain the variation in one of these performance variables. Fit a full model (using all of your selected variables). Remove variables that do not add information to the model. The goal is to explain as much as possible of the performance.

You might use p-values, $R^2$ or the `AIC()` function to select variables. 

Check your assumptions and interpret the final model. 

For this assignment you are expected to use the course literature to motivate your work-flow.


Diez, D. M., C. D. Barr, and M. Çetinkaya-Rundel. 2015. [OpenIntro Statistics. OpenIntro, Incorporated](https://leanpub.com/openintro-statistics) 

Navarro, Daniel. 2019. [Learning Statistics with R: A Tutorial for Psychology Students and Other Beginners (Version 0.6.1)](https://learningstatisticswithr.com/book/)



