---
title: "Assignment Week 4 Soultions/Examples"
output: html_document
---

## Loading data for the assignment and creating descriptive graphics:




```{r load_data_figure, message=FALSE, warning = FALSE}
library(tidyverse) # Tidyverse includes the readr package

read_csv("./data/strengthTests.csv") %>%
        filter(!is.na(load)) %>% # Removes missing values
        # Summarise values per group, timepoint, exercise
        group_by(timepoint, group, exercise) %>% 
        summarise(m = mean(load), 
                  s = sd(load), 
                  n = n()) %>%
        group_by() %>% # removes grouping from data frame
        # Change levels in timepoint
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", 
                                             "session1", 
                                             "post"))) %>%
        # Creates plot
        ggplot(aes(timepoint, m, group = group, color = group)) + 
        geom_errorbar(aes(ymin = m-s, ymax = m + s), 
                      width = 0.2, 
                      position = position_dodge(width = 0.2)) +
        geom_line(position = position_dodge(width = 0.2)) + 
        facet_wrap(~ exercise, scales = "free")

```


A pre- to post-training design may be analyzed by post-values only. This is "ineffective" as we need many more participants to show a difference that is true. The example below is an example of such analysis. 


```{r, message=FALSE, warning = FALSE}

## Bad example, inefficient statistical test
temp <- read_csv("./data/strengthTests.csv") %>%
        filter(!is.na(load), 
               timepoint == "post", 
               exercise == "isom") %>%
        print()
 

t.test(load ~ group, data = temp)


```

Instead of using post-training data only, we can model the change. This can also be done in a t-test:

```{r, message=FALSE, warning = FALSE}

change.data <- read_csv("./data/strengthTests.csv") %>%
        filter(!is.na(load), 
               exercise == "isom") %>%
        mutate(timepoint = if_else(timepoint == "session1", 
                                   "pre", timepoint)) %>%
        group_by(subject, timepoint, group, sex) %>%
        summarise(load = mean(load, na.rm = TRUE)) %>%
        
        pivot_wider(names_from = timepoint, 
                    values_from = load) %>%
        mutate(change = (post - pre)/pre * 100) %>%
        ungroup() %>%
        mutate(pre.mc = pre - mean(pre)) 


t.test(change ~ group, data = change.data, var.equal = TRUE)

```

The t-test "controls" for baseline as we model the change from each individual baseline. To further account for the baseline, we can include it in a model.
The t-test can be performed as a linear model, which can be extended.

```{r, message=FALSE, warning = FALSE}

## Linear model (same as t-test)
m1 <- lm(change ~ group, data = change.data)
summary(m1)


## Linear model with baseline values as covariate
m2 <- lm(change ~ pre.mc + group, data = change.data)
summary(m2)





```

### Creating tables

In the code chunk options, remember to set `results = "asis"` to get a table from `kable`.

```{r table1, results = "asis", message=FALSE, warning = FALSE}

library(broom) # is used for formatting model summaries 
library(knitr) # is used for making tables

# A simple table from m2 above

tidy(m2) %>%
        # Change names of the term column
        mutate(term = if_else(term == "(Intercept)", "Intercept", 
                              if_else(term == "pre.mc", "Baseline",
                                      "Single-set"))) %>%
        # kable is used to "create" the table. In practice, it 
        # translates your table to html-code.
        kable(col.names = c("Parameter", 
                            "Estimate", "SE", 
                            "t-value", "p-value"), 
              digits = c(NA, 2, 2, 2, 3))


```


## Mixed models


```{r mixed-models, message=FALSE, warning = FALSE}

library(lme4); library(broom)

# Alternative 1 (Two pre- time-points)
raw.data <- read_csv("./data/strengthTests.csv") %>%
        filter(!is.na(load), 
               exercise == "isom") %>%
        mutate(timepoint = if_else(timepoint == "session1", 
                                   "pre", timepoint)) %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre", "post"))) %>%
        
        print()

# Alternative 2 (combine pre and session1 into one time-point)
raw.data <- read_csv("./data/strengthTests.csv") %>%
        filter(!is.na(load), 
               exercise == "isom") %>%
        mutate(timepoint = factor(timepoint, 
                                  levels = c("pre",
                                             "session1", 
                                             "post"))) %>%
        
        print()



### Mixed model
# See lessons for interpretation of the model
m2 <- lmer(load ~ timepoint * group + (1|subject), 
           data = raw.data)

# Model summary
summary(m2)

# Diagnostics
plot(m2)

# Confidence intervals
confint(m2)


```


```{r table, results = "asis", message=FALSE, warning = FALSE}

library(knitr); library(broom)


# creating a table

# We can combine two data frames by adding columns or rows

# The cbind function adds two data frames with simialr rows.

# First we save a tidy output of model 2
 
tidy.m2 <- tidy(m2) %>%
        filter(group == "fixed") # removes random effects from summary

# Next we store the confidence intervals from confint
# We have to select row 3 to 8 to get the same number of rows
confint.m2 <- data.frame(confint(m2))[c(3:8),]

# Next we can comine these by column
cbind(tidy.m2, confint.m2) %>%
        select(-group) %>% # deselect "group"-column
        kable(row.names = FALSE, 
              col.names = c("Parameter", 
                            "Estimate", 
                            "SE", 
                            "Statistic", 
                            "Lower CI", 
                            "Upper CI"), 
              digits = c(NA, 1, 1, 2, 1, 1)) %>%
                      print()




```








