---
title: "Week 2 Solutions"
author: "Daniel Hammarström"
date: '2019-10-21'
output:
  word_document: default
  html_document: default
---

## Problem 1

```{r problem1, warning = FALSE, message=FALSE, echo=FALSE}
library(tidyverse); library(readxl)

cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, timepoint, weight.T1, VO2.max) %>%
        mutate(vo2max.kg = VO2.max / weight.T1) %>%
        select(-weight.T1, -VO2.max) %>%
        filter(timepoint %in% c("pre", "meso1")) %>%
        pivot_wider(names_from = timepoint, 
                    values_from = vo2max.kg) 


# descriptive statistics
pre.m <- round(mean(cdat$pre), 1)
meso1.m <- round(mean(cdat$meso1), 1)

pre.sd <- round(sd(cdat$pre), 1)
meso1.sd <- round(sd(cdat$meso1), 1)

# Creating the t-test
ttest_vo2max <- t.test(cdat$meso1, cdat$pre, paired = TRUE)

est <- round(ttest_vo2max$estimate, 1)
conf <- round(ttest_vo2max$conf.int, 2)
p.value <- round(ttest_vo2max$p.value,3)


```

VO~2max~ ml kg^-1^ min^-1^ increased from pre (mean (SD), `r pre.m` (`r pre.sd`)) to meso1 (`r meso1.m` (`r meso1.sd`)) with mean difference being `r est` ml kg^-1^ min^-1^ (95% CI: `r conf[1]`, `r conf[2]`, p=`r p.value`). The null-hypothesis of no differences between means could not be rejected at the $\alpha$ = 0.05 level. Instead of a paired-sample t-test, we could calculate difference between time-point for each participant and use these data in a one sample t-test. 


## Problem 2
```{r problem2, echo = FALSE, warning= FALSE, message=FALSE}


cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, timepoint, sj.max) %>%
        pivot_wider(names_from = timepoint, 
                    values_from = sj.max) %>%
    mutate(change = 100 * (meso1 - pre) / pre) 


sj_ttest <- t.test(cdat$change, mu = 0)

est <- round(sj_ttest$estimate, 2)
conf <- round(sj_ttest$conf.int, 2)
```

The mean difference was estimated to `r est`% 95% CI: [`r conf[1]`, `r conf[2]`]. We would reject the null-hypothesis at the level of $\alpha$ = 0.05. 

## Problem 3

```{r problem3, echo = FALSE, warning=FALSE, message=FALSE}

cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, group, timepoint, VO2.max) %>%
        filter(group != "MIX") %>%
        pivot_wider(names_from = timepoint, 
                    values_from = VO2.max) %>%
        mutate(change = 100 * (meso3-pre)/pre) 


sum_stats <- cdat %>%
        group_by(group) %>%
        summarise(m = round(mean(change, na.rm = TRUE), 2),
                  s = round(sd(change, na.rm = TRUE), 2)) 


ttest_group <- t.test(change ~ group,  data = cdat)


est <- round(ttest_group$estimate[1] - ttest_group$estimate[2], 2)
conf <- round(ttest_group$conf.int, 2)
pval <- round(ttest_group$p.value, 3)

```

The mean change score in DECR was `r sum_stats[1,2]`% (SD: `r sum_stats[1, 3]`) and `r sum_stats[2,2]` (`r sum_stats[2, 3]`) in INCR. The mean difference was `r est`%-point (95% CI: [`r conf[1]`, `r conf[2]`], p-value=`r pval`). The null hypothesis of no difference between groups was not rejected at $\alpha$ = 0.05.

## Code for each problem

### Problem 1

```{r problem1_code, eval = FALSE, echo = TRUE}
library(tidyverse); library(readxl)

cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, timepoint, weight.T1, VO2.max) %>%
        mutate(vo2max.kg = VO2.max / weight.T1) %>%
        select(-weight.T1, -VO2.max) %>%
        filter(timepoint %in% c("pre", "meso1")) %>%
        pivot_wider(names_from = timepoint, 
                    values_from = vo2max.kg) 


# descriptive statistics
pre.m <- round(mean(cdat$pre), 1)
meso1.m <- round(mean(cdat$meso1), 1)

pre.sd <- round(sd(cdat$pre), 1)
meso1.sd <- round(sd(cdat$meso1), 1)

# Creating the t-test
ttest_vo2max <- t.test(cdat$meso1, cdat$pre, paired = TRUE)

est <- round(ttest_vo2max$estimate, 1)
conf <- round(ttest_vo2max$conf.int, 2)
p.value <- round(ttest_vo2max$p.value,3)


```

### Problem 2

```{r problem2_code, echo = TRUE, eval = FALSE}


cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, timepoint, sj.max) %>%
        pivot_wider(names_from = timepoint, 
                    values_from = sj.max) %>%
    mutate(change = 100 * (meso1 - pre) / pre) 


sj_ttest <- t.test(cdat$change, mu = 0)

est <- round(sj_ttest$estimate, 2)
conf <- round(sj_ttest$conf.int, 2)
```

### Problem 3

```{r problem3_code, echo = TRUE, eval = FALSE}

cdat <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>%
        select(subject, group, timepoint, VO2.max) %>%
        filter(group != "MIX") %>%
        pivot_wider(names_from = timepoint, 
                    values_from = VO2.max) %>%
        mutate(change = 100 * (meso3-pre)/pre) 


sum_stats <- cdat %>%
        group_by(group) %>%
        summarise(m = round(mean(change, na.rm = TRUE), 2),
                  s = round(sd(change, na.rm = TRUE), 2)) 


ttest_group <- t.test(change ~ group,  data = cdat)


est <- round(ttest_group$estimate[1] - ttest_group$estimate[2], 2)
conf <- round(ttest_group$conf.int, 2)
pval <- round(ttest_group$p.value, 3)

```



