---
title: "Exercises Week 1"
---

[Go to basic R](ex1_intro.html)

[Go to summarising data](ex1_summarise.html)

[Go to making figures](ex1_import.html)


 <script language="javascript"> 
    function toggle(num) {
      var ele = document.getElementById("toggleText" + num);
      var text = document.getElementById("displayText" + num);
      if(ele.style.display == "block") {
        ele.style.display = "none";
        text.innerHTML = "show";
      }
      else {
        ele.style.display = "block";
        text.innerHTML = "hide";
      }
   } 
  </script>

**Exercise 1.** Using the cycling study data set, plot the $\dot VO_{2max}$ as a function of time (`timepoint` of x-axis and `VO2.max` on the y-axis). Plot box-plots per group.

  <a id="displayText" href="javascript:toggle(1);">Solution Ex. 1</a>
  <div id="toggleText1" style="display: none">

```{r, eval = FALSE}
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        select(subject, group, timepoint, VO2.max) %>%
        ggplot(aes(timepoint, VO2.max, color = group)) + geom_boxplot()

```

  </div>
  <br>


**Exercise 2.** Using the cycling study data set, plot the $\dot VO_{2max}$ per kg body weight `weight.T1`as a function of time (`timepoint` of x-axis and `VO2.max` on the y-axis). Plot box-plots per group. Control the order of the `timepoint` factor so that "pre" is first followed by "meso1", "meso2" and "meso3".

  <a id="displayText" href="javascript:toggle(2);">Solution Ex. 2</a>
  <div id="toggleText2" style="display: none">

```{r, eval = FALSE}
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        select(subject, group, timepoint, VO2.max, weight.T1) %>%
        mutate(vo2max.kg = VO2.max / weight.T1,
               timepoint = factor(timepoint, levels = c("pre", "meso1", "meso2", "meso3"))) %>%
        
        ggplot(aes(timepoint, VO2.max, color = group)) + geom_boxplot()

```

  </div>
  <br>


**Exercise 3.** Using the cycling study data set, plot the $\dot VO_{2max}$ per kg body weight `weight.T1`as a function of time (`timepoint` of x-axis and `VO2.max` on the y-axis). Instead of box plots, set color per group but show each data point with points and connected line. The lines should connect per subject, you need to control `group =` in `aes`. Control the order of the `timepoint` factor so that "pre" is first followed by "meso1", "meso2" and "meso3".

  <a id="displayText" href="javascript:toggle(3);">Solution Ex. 3</a>
  <div id="toggleText3" style="display: none">

```{r, eval = FALSE}
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        select(subject, group, timepoint, VO2.max, weight.T1) %>%
        mutate(vo2max.kg = VO2.max / weight.T1,
               timepoint = factor(timepoint, levels = c("pre", "meso1", "meso2", "meso3"))) %>%
        ggplot(aes(timepoint, VO2.max, color = group, group = subject)) + geom_point() + geom_line()

```

  </div>
  <br>


**Exercise 4.** Do your best to replicate the figure below. 

Things to google: 

- "remove legend from ggplot"
- "Adding plus minus sign to ggplot title"
- "Change shape of geom_point ggplot2"



```{r, message = FALSE, warnings = FALSE, echo = FALSE}
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        mutate(vo2max.kg = VO2.max / weight.T1) %>%
        select(subject, group, timepoint, vo2max.kg) %>%
        pivot_wider(names_from = timepoint, 
                    values_from = vo2max.kg) %>%
        mutate(change = ((meso3 / pre)-1) * 100) %>%
        group_by(group) %>%
        summarise(m = mean(change, na.rm = TRUE), 
                  s = sd(change, na.rm = TRUE)) %>%
        ggplot(aes(group, m, fill = group)) + 
        geom_errorbar(aes(ymin = m - s, ymax = m + s), width = 0.2) +
        geom_point(size = 4, shape = 21) +
        labs(x = "Group", y = "VO2max per kg % change pre to meso3 (mean \U00B1 SD)") +
        theme(legend.position = "none")
 
```



  <a id="displayText" href="javascript:toggle(4);">Solution Ex. 4</a>
  <div id="toggleText4" style="display: none">

```{r, eval = FALSE}
# load packages
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        # Calculate relative vo2max
        mutate(vo2max.kg = VO2.max / weight.T1) %>%
        # Select variables
        select(subject, group, timepoint, vo2max.kg) %>%
        # Make the data into wide format
        pivot_wider(names_from = timepoint, 
                    values_from = vo2max.kg) %>%
        # Calculate percentage change pre to meso3
        mutate(change = ((meso3 / pre)-1) * 100) %>%
        # group by group and calculate statistics
        group_by(group) %>%
        summarise(m = mean(change, na.rm = TRUE), 
                  s = sd(change, na.rm = TRUE)) %>%
        # Using fill per group 
        ggplot(aes(group, m, fill = group)) + 
        # Adding error bar first, the overplot with points
        geom_errorbar(aes(ymin = m - s, ymax = m + s), width = 0.2) +
        # shape = 21 gives circles that can be filled
        geom_point(size = 4, shape = 21) +
        labs(x = "Group", y = "VO2max per kg % change pre to meso3 (mean \U00B1 SD)") +
        theme(legend.position = "none") # legend position = "none" removes legend.

```

  </div>
  <br>






