---
title: "Exercises Week 1"
---

[Go to basic R](ex1_intro.html)

[Go to summarising data](ex1_summarise.html)

[Go to making figures](ex1_figures.html)

 <script language="javascript"> 
    function toggle(num) {
      var ele = document.getElementById("toggleText" + num);
      var text = document.getElementById("displayText" + num);
      if(ele.style.display == "block") {
        ele.style.display = "none";
        text.innerHTML = "show";
      }
      else {
        ele.style.display = "block";
        text.innerHTML = "hide";
      }
   } 
  </script>

**Exercise 1.** Import and *Select* `subject`, `timepoint` and `sj.max` from the cycling study data set. Print your results. What does printing do?

  <a id="displayText" href="javascript:toggle(1);">Solution Ex. 1</a>
  <div id="toggleText1" style="display: none">

```{r, eval = FALSE}
# load packages
library(tidyverse)
library(readxl)

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") # remember to use the na argument

cyclingStudy %>%
        select(subject, timepoint, sj.max) %>%
        print()


```

Alternatively, everything can be written in one pipe:

```{r, eval = FALSE}

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
 select(subject, timepoint, sj.max) %>%
        print()

```


  </div>
  </br>  
  
**Exercise 2.** Filter the cycling data set based on time-point keep only observations from time-point pre. 


  <a id="displayText" href="javascript:toggle(2);">Solution Ex. 2</a>
  <div id="toggleText2" style="display: none">
<br> 

```{r, eval = FALSE}

cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
 filter(timepoint == "pre") %>%
        print()

```


  </div>
  </br>  
  

**Exercise 3.** Select only columns `subject` and `sj.max`. Filter the cycling data set based on time-point keep only observations from time-point pre and meso1. . In what order must you do the operations in order to get it to work?


  <a id="displayText" href="javascript:toggle(3);">Solution Ex. 3</a>
  <div id="toggleText3" style="display: none">
<br> 

```{r, eval = FALSE}
cyclingStudy <- read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
        filter(timepoint %in% c("pre", "meso1")) %>%
        select(subject, sj.max) %>%
        print()
```


  </div>
  </br>  
  

**Exercise 4.** `weight.T1` is the weight in kg and `VO2.max` is the $\dot VO_{2max}$ in $ml \times min^{-1}$ from each test. Select the needed variables, filter needed observations, create a new variable $\dot VO_{2max} / kg$ and calculate the percentage change from time-point "pre" to "meso3". This operation should include `mutate()`, `filter()`, `pivot_wider`.


  <a id="displayText" href="javascript:toggle(4);">Solution Ex. 4</a>
  <div id="toggleText4" style="display: none">
<br> 

```{r, eval = FALSE}
read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
        filter(timepoint %in% c("pre", "meso3")) %>%
        select(subject, timepoint, VO2.max, weight.T1) %>%
        mutate(relative.vo2max = VO2.max / weight.T1) %>%
        select(subject, timepoint, relative.vo2max) %>%
        pivot_wider(names_from = timepoint, values_from = relative.vo2max) %>%
        mutate(change = ((meso3 / pre)-1)*100) %>%
        print()
```


  </div>
  </br>  
  

**Exercise 5.** Do the same as above but keep the `group` variable. Calculate the mean of each group in percentage change from pre- to meso3. In the data set there are three groups: "INCR" increased intensity through the training period, "DECR" decreased intensity and "MIX" performed a mix of intensities each week.


  <a id="displayText" href="javascript:toggle(5);">Solution Ex. 5</a>
  <div id="toggleText5" style="display: none">
<br> 

```{r, eval = FALSE}
read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
        filter(timepoint %in% c("pre", "meso3")) %>%
        select(subject, group, timepoint, VO2.max, weight.T1) %>%
        mutate(relative.vo2max = VO2.max / weight.T1) %>%
        select(subject, group, timepoint, relative.vo2max) %>%
        pivot_wider(names_from = timepoint, values_from = relative.vo2max) %>%
        mutate(change = ((meso3 / pre)-1)*100) %>%
        group_by(group) %>%
        summarise(m = mean(change, na.rm = TRUE)) %>%
        print()
```

The `n()`function can be used with `summarise()`. `n()` returns the number of observations including NAs. We can include `n()` to count the number of observations. If we are only interested in values that are not NA we could use the `sum()` function and count the number of values that are not NA. `is.na()` says TRUE if it finds a NA. if we inverse it, we will get a counter of non NAs

<br>


```{r, eval = FALSE}
read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
        filter(timepoint %in% c("pre", "meso3")) %>%
        select(subject, group, timepoint, VO2.max, weight.T1) %>%
        mutate(relative.vo2max = VO2.max / weight.T1) %>%
        select(subject, group, timepoint, relative.vo2max) %>%
        pivot_wider(names_from = timepoint, values_from = relative.vo2max) %>%
        mutate(change = ((meso3 / pre)-1)*100) %>%
        group_by(group) %>%
        summarise(m = mean(change, na.rm = TRUE), 
                  n = n(), 
                  non.na = sum(!is.na(change))) %>%
        print()
```



  </div>
  </br>  
  


**Exercise 6.** $\dot VO_2$ meassures were collected during submaximal exercise, variables `VO2.125:VO2.375` are $\dot VO_2$ measurements from each watt. Transform the data to get this to tidy format.


  <a id="displayText" href="javascript:toggle(6);">Solution Ex. 6</a>
  <div id="toggleText6" style="display: none">
<br> 

```{r, eval = FALSE}
read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
       select(subject, timepoint, VO2.125:VO2.375) %>%
        pivot_longer(names_to = "watt", 
                     names_prefix = "VO2.",
                     names_ptypes = list(watt = numeric()),
                     values_to = "VO2", 
                     cols = VO2.125:VO2.375) %>%
        print()
```


  </div>
  </br> 


**Exercise 7.** $\dot VO_2$ meassures were collected during submaximal exercise, variables `VO2.125:VO2.375` are $\dot VO_2$ measurements from each watt. Transform the data to get this to tidy format. Calculate the mean $\dot VO_2$ for each group, timepoint and watt.


  <a id="displayText" href="javascript:toggle(7);">Solution Ex. 7</a>
  <div id="toggleText7" style="display: none">
<br> 

```{r, eval = FALSE}
read_excel("./data/cyclingStudy.xlsx", na = "NA") %>% # remember to use the na argument
       select(subject, group, timepoint, VO2.125:VO2.375) %>%
        pivot_longer(names_to = "watt", 
                     names_prefix = "VO2.",
                     names_ptypes = list(watt = numeric()),
                     values_to = "VO2", 
                     cols = VO2.125:VO2.375) %>%
        group_by(group, timepoint, watt) %>%
        summarise(vo2 = mean(VO2, na.rm = TRUE)) %>%
        print()

```


  </div>
  </br> 









