---
title: "Pre- to post-treatment analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The 10 vs. 30RM study

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(readxl); library(tidyverse)
n <- read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        summarise(n =  n_distinct(subject)) %>%
        as.numeric()
```

In this study, n = `r n` participants were renadomized to completed a strength training intervention using loads corresponding to either 30 or 10 repetition maximum. Strength was evaluated as one repetition maximum (1RM) in the benchpress, legpress and bicepscurl. We will analyze strength test data from this trial in the coming week. 

### Explorative analysis

Use descriptive methods (summary statistics and figures to describe results from the trial). What are the mean and standard deviations of the `load` variable for each `timepoint`, `exercise` and `group`. Use tables and figures to show the results.


 <script language="javascript"> 
    function toggle(num) {
      var ele = document.getElementById("toggleText" + num);
      var text = document.getElementById("displayText" + num);
      if(ele.style.display == "block") {
        ele.style.display = "none";
        text.innerHTML = "show";
      }
      else {
        ele.style.display = "block";
        text.innerHTML = "hide";
      }
   } 
  </script>


 <a id="displayText" href="javascript:toggle(1);">Here is a possible solution for a table</a>
  <div id="toggleText1" style="display: none">

We want to summarise data per group, exercise and time-point and we want the means, standad deviations and the number of observations. 

```{r, eval = FALSE}


ten_vs_thirty_summary <- read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        group_by(exercise, timepoint, group) %>%
        summarise(Mean = mean(load, na.rm = TRUE), 
                  SD = sd(load, na.rm = TRUE), 
                  n = n()) 


# To create a table we need the kable function from the knitr package

library(knitr)

ten_vs_thirty_summary %>%
kable(digits = c(NA, NA, NA, 1, 1, 0), 
      col.names = c("Exercise", "Time-point", "Group", "Mean", "SD", "n"))


```


  </div>
  </br>  



 <a id="displayText" href="javascript:toggle(2);">Here is a possible solution for a figure</a>
  <div id="toggleText2" style="display: none">

Using the same summary statistics, we can create a figure. We might want to sort variables so that time-points are in the right order. In the figure we can use geom_errorbar to display SD


```{r, eval = FALSE}

# create a position for all variables and store in object
pos <- position_dodge(width = 0.2)


ten_vs_thirty_summary %>%
        ungroup() %>% # Needed to mutate the timepoint variable
        mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
        ggplot(aes(timepoint, Mean, color = group, group = group)) + 
        geom_errorbar(aes(ymin = Mean - SD, ymax = Mean + SD), position = pos) +
        geom_line(position = pos) +
        geom_point(shape = 21, position = pos) + 
                facet_wrap(~exercise, scales = "free")
        
```



  </div>
  </br>  


What can you say about the effect of training with 10RM vs. 30RM on muscle strength based on these observations?


### Change from baseline

We will now focuse on the change from baseline (`pre`) to the `mid` time-point. To only use this data, use the filter function:

```{r, eval = FALSE}

ten_vs_thirty <- read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        filter(timepoint != "post") %>%
        print()

```

Calculate the average percentage change from baseline in each group and create a graph of the results. Make a plan of your code before you start writing it!


 <a id="displayText" href="javascript:toggle(3);">Here is a possible solution for a figure</a>
  <div id="toggleText3" style="display: none">

Here we want to perform the following steps: 

1. Filter the data to only contain pre and mid time-points
2. Make the data set wider and calculate percentage change as $\frac{mid-pre}{pre} \times 100$
3. Group the data set by group and exercise
4. Summarise the data and create a figure with groups on the x-axis and %-change on the y-axis.

 <a id="displayText" href="javascript:toggle(4);">Example code</a>
  <div id="toggleText4" style="display: none">

```{r, eval = FALSE}

read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        filter(timepoint != "post") %>%
        pivot_wider(names_from = timepoint, 
                    values_from = load) %>%
        mutate(change = (mid-pre) / pre * 100) %>%
        group_by(group, exercise) %>%
        summarise(m = mean(change, na.rm = TRUE), 
                  s = sd(change, na.rm = TRUE)) %>%
        ggplot(aes(group, m)) + 
        geom_errorbar(aes(ymin = m - s, ymax = m + s)) +
        geom_point() + 
        facet_wrap(~ exercise)

```



  </div>
  </br>  

  </div>
  </br>  

### Model the change

Focus on the benchpress data. State a hypothesis that compares the two groups and create a test that test against the corresponding null-hypothesis. Remember that we did this using both `t.test()` and `lm()`. Using `var.equal = TRUE` do the corresponding test also in `lm`. Write a short summary of your results!


 <a id="displayText" href="javascript:toggle(5);">Example code</a>
  <div id="toggleText5" style="display: none">

```{r, eval = FALSE}

tt_data <- read_excel("./data/ten_vs_thirty.xlsx", na = "NA") %>%
        filter(timepoint != "post") %>%
        pivot_wider(names_from = timepoint, 
                    values_from = load) %>%
        mutate(change = (mid-pre) / pre * 100) %>%
        filter(exercise == "benchpress") %>%
        print()

## The t.test

t.test(change ~ group, data = tt_data, var.equal = TRUE)


## The linear model

summary(lm(change ~ group, data = tt_data))

```






  </div>
  </br>  


### Accounting for the baseline -- Regression towards the mean

### Accounting for the baseline -- Differences between group and mixed models

### Multiple time-point -- Mixed models











