---
title: "Importing data"
---

## Resources

[Chapter 11]("https://r4ds.had.co.nz/data-import.html") in R for data science describes data import of .csv. The [readxl]("https://readxl.tidyverse.org/) package is used for reading excel files. The [haven]("https://haven.tidyverse.org/") package can be used to import files from SAS, SPSS and STATA /(we will worry about that later).

## Learning objectives

Get familiar with importing data from 
- Excel files
- Flat files (.csv, .tsv)
- And know about importing from other filetypes


## Data sets used in the course
We will primarily use four data sets in the course. Below is a table of where they can be found and publications connected to the data. The data sets will be used for different purposes. 

|Data set      |Author/Year      |Public|Publication (DOI)           |Download                                       |Filetype |
|--------------|-----------------|------|----------------------------|-----------------------------------------------|---------|
|Hypertrophy   |Haun et al. 2019 |Yes   |10.3389/fphys.2019.00297    |https://ndownloader.figshare.com/files/14702420|.csv     |
|Cycling       |Sylta et al. 2016|No    |10.1249/MSS.0000000000001007|Canvas                                         |.xlsx    |
|Diet          |Hall et al. 2019 |Yes   |10.1016/j.cmet.2019.05.008  |https://osf.io/jh53q/ (multiple files)         |.sas7bdat| 
|Ten vs. Thirty|Øfsteng et al.   |No    |NA                          |Canvas                                         |.xlsx    |

## Importing excel-files
Importing excel-files is straight forward as long as they are nicely formatted. A bad format is when the data is not tabular.

To import excel files (.xls, and .xlsx) we need the package `readxl`. The package contains the function `read_excel()`. Install the package and access the help-file for the function `read_excel()`.

We will start by importing a data set based on part of a cycling training study. This data set has to be downloaded manually from Canvas. Go to data Lectures/Data, there you will find a link to the `cyclingStudy.xlsx` file. Download the file and put it in your projects data folder.

Importing the file could look like this:

```{r, eval = FALSE}
library(readxl)
cyclingStudy <- read_excel("./data/cyclingStudy")
```

We "store" the file in the environment in an object called `cyclingStudy`. 

Important arguments in the `read_excel()` function includes:

- `path` which is the location of your file
- `sheet` if you are working with a excel-file containing multiple sheets
- `na` specifies what are missing data, this is important when you import numeric data that contains character string data to show e.g. missing values
- `skip` tells `read_excel` how many rows to skip before reading the data. This can be helpful when the tabular part of the data does not start at row 1.

There are more arguments, use `?read_excel` to find out.

## Importing .csv files
A csv-file (comma separated values) is a text-file type file. It could not be simpler! Important to know about them is that column is separated by a comma, semicolon or similar. If the file is comma separated we can simply use the `read_csv()` function from the `readr` package. 

The .csv file we will be using during this course is available online. We can without so much fuzz load it directly using

```{r, eval = FALSE}
library(readr)
hypertrophy <- read_csv("https://ndownloader.figshare.com/files/14702420")
```

When downloading files from the internet, we save the file first using the `download.file()` function. This function should be available in your R session.
Let's store the file in the data folder:

```{r, eval = FALSE}
download.file(url = "https://ndownloader.figshare.com/files/14702420", destfile = "./data/hypertrophy.csv")
```

We can now access the file from our data folder.

## Importing other file formats
We will be analyzing a [diet study](https://www.cell.com/cell-metabolism/fulltext/S1550-4131(19)30248-7) in the course. This study has all the data and code (SAS) [published](https://osf.io/rx6vm/). However, it is zipped and needs manual download. We will use the [haven]("https://haven.tidyverse.org/") package to get the data into R.  






