---
title: "Corrrelation and Linear regression"
author: "Daniel Hammarström"
date: "2019-10-21"
output: beamer_presentation
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---

# Association between variables

* A measure of association between continuous variables is the correlation (Pearson's correlation coefficient).

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height=4, fig.width=7, fig.align='center'}
library(tidyverse); library(cowplot)
set.seed(2)

N <- 100 # number of pairs in simulation

# Create some randomly distributed X values representing some quantrity

X <- runif(N, 100, 5000) / 1000

# generate some y values according to a linear model. Here, they have
# slope = 1.2, intercept = 1 and Gaussian (normal) random noise added with
# mean = 0 and standard deviation = 1. 


Y1 <- 1*X + 1 + rnorm(N,0,1)
Y0 <- 0*X + 1 + rnorm(N,0,1)
Y_1 <- -1*X + 1 + rnorm(N,0,1)



df <- data.frame(x = X, "Positive"= Y1, "No correlation" = Y0, "Negative" = Y_1) %>%
        gather(y, value, c(Positive, No.correlation, Negative)) 

sum <- df %>%
        group_by(y) %>%
        summarise(corr = cor(x, value)) %>%
        mutate(x = c(2.5, 2.5, 2.5),
               value = c(3, 5,-0.5)) %>%
        mutate(y = factor(y, levels = c("Positive", "No.correlation", "Negative"),
                          labels = c("Positive\ncorrelation", "No Correlation", "Negative\ncorrelation")))


df %>%
        mutate(y = factor(y, levels = c("Positive", "No.correlation", "Negative"),
                          labels = c("Positive\ncorrelation", "No Correlation", "Negative\ncorrelation"))) %>%
        ggplot(aes(x, value)) + geom_point() +
        facet_grid(.~y)+ theme_minimal() +  
        theme(panel.border = element_rect(color = "black",                                                                     fill = NA)) + 
        geom_text(data = sum, aes(x, value, label = paste0("Correlation = ",round(corr, 3)))) + 
        geom_smooth(method = "lm", se = FALSE) + 
        xlab("Variable A") + ylab("Variable B")


```

# Correlation gives a unitless "strength of association"

* Estimates of association ($r$) is limited to $-1 \leq r \leq 1$.
* When r approaches $\pm 1$, the association is stronger, estimates close to 0 suggest no association.

```{r,echo = FALSE, warning = FALSE, message = FALSE, fig.height=4, fig.width=7, fig.align='center'}
set.seed(2)

N <- 100 # number of pairs in simulation

# Create some randomly distributed X values representing some quantrity

X <- runif(N, 100, 5000) / 1000

# generate some y values according to a linear model. Here, they have
# slope = 1.2, intercept = 1 and Gaussian (normal) random noise added with
# mean = 0 and standard deviation = 1. 


Y1 <- 1*X + 1 + rnorm(N,0,1)
Y2 <- 1*X + 1 + rnorm(N,0,4)
Y3 <- 1*X + 1 + rnorm(N,0,10)



df <- data.frame(x = X, Y1, Y2, Y3) %>%
        gather(y, value, Y1:Y3) 

sum <- df %>%
        group_by(y) %>%
        summarise(corr = cor(x, value)) %>%
        mutate(x = c(2.5, 2.5, 2.5),
               value = c(30, 30, 30)) 

df %>%
        ggplot(aes(x, value)) + geom_point() +
        facet_grid(.~y)+ theme_minimal() +  
        theme(panel.border = element_rect(color = "black",                                                                     fill = NA),
              strip.text.x = element_blank())  +
        geom_smooth(method = "lm", se = FALSE) +
        geom_text(data = sum, aes(x = x, y = value, label = paste("r = ", sprintf("%.3f",corr))))
```

# Assumptions in correlation

* Continuous variables, paired observations
* Bivariate normal distribution(?) -- Both variables should be bell shaped.
* Linear relationship between variables
* Be careful when there are outliers, examine the effect of extreme data points.


# Correlation in R

```{r, eval = FALSE}
Yj <- c(25.2, 26.9, 21.7, 15.8, 
        26.0, 20.4, 18.5, 15.5, 15.6, 16.0)

Yk <- c(21.9, 25.7, 23.6, 29.6, 
        24.9, 23.4, 23.5, 25.1, 24.0, 21.5)
 
# The Pearson's product moment 
# correlation coefficient
cor(Yk, Yj, method = "pearson")   

# The Pearson's product moment 
# correlation coefficient with 
# test statistic
cor.test(Yk, Yj, method = "pearson")   
```

# Regression models the relationship between variables 

* The regression model describe more aspects of the relationship between variables than the correlation.
* The equation for the straight line:

$$y = mx+c ~~~~~~~~ y = {slope} \times x + {intercept}$$

```{r, echo = FALSE, warning = FALSE, message = FALSE, out.width = "60%",out.height = "50%",fig.width = 4, fig.height=4, fig.align="center"}

ggplot(data = data.frame(x = c(0, 5), y = c(0, 5)), aes(x, y)) + geom_point(size = 0.1) +
        geom_hline(yintercept = 0, color = "gray60", size = 1) + 
        geom_vline(xintercept = 0, color = "gray60", size = 1) +
        geom_abline(slope = 1, intercept = 2, size = 1, lty = 2) + 
        annotate(geom = "text", 
                 label = "Intercept", 
                 x = 0.5, y = 2, parse = TRUE, 
                 angle = 0) +
         annotate(geom = "text", 
                 label = "Slope", 
                 x = 0.5, y = 2.7, parse = TRUE, 
                 angle = 50) +
                theme_minimal() + scale_x_continuous(limits = c(-1, 3), expand = c(0,0)) +
        scale_y_continuous(limits = c(0, 3), expand = c(0,0)) +
        xlab("y") + ylab("x")
        

```

# Regression estimates the line that best fits the data

* The basic univariate regression model

$$y = \beta_0 + \beta_1x + \epsilon$$

* $\beta_0$ is the model intercept (or constant)
* $\beta_1$ is the slope of the straight line
* $\epsilon$ is the unexplained error
* Model parameters ($\beta_0, \beta_1$) are estimated using sample data


# Interpret slopes and intercepts

```{r,echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height=4}
a <- ggplot(data = data.frame(x = c(0, 5), y = c(0, 5)), aes(x, y)) + geom_point() +
        geom_abline(slope = 1, intercept = 2, size = 1, lty = 2) + 
        geom_abline(slope = -1, intercept = 2, size = 1, lty = 3) + 
        geom_abline(slope = 0, intercept = 2, size = 1, lty = 4) + 
        annotate(geom = "text", 
                 label = paste(expression(beta[1]==1)), 
                 x = 1, y = 3.2, parse = TRUE, 
                 angle = 60) +
        annotate(geom = "text", 
                 label = paste(expression(beta[1]==0)), 
                 x = 1, y = 1.2, parse = TRUE, 
                 angle = -60) +
          annotate(geom = "text", 
                 label = paste(expression(beta[1]==0)), 
                 x = 1, y = 2.2, parse = TRUE) +
        annotate(geom = "text", 
                 label = paste(expression(beta[0] == 2)), 
                 x = 0.5, y = 0.6, parse = TRUE) +
        theme_minimal() + scale_x_continuous(limits = c(0, 4), expand = c(0,0)) +
        scale_y_continuous(limits = c(0, 4), expand = c(0,0)) +
        xlab("") + 
        theme(panel.background = element_rect(color = "black"))

b <- ggplot(data = data.frame(x = c(0, 5), y = c(0, 5)), aes(x, y)) + geom_point() +
        geom_abline(slope = 1, intercept = 2, size = 1, lty = 2) + 
        geom_abline(slope = 1, intercept = 1, size = 1, lty = 3) + 
        geom_abline(slope = 1, intercept = 0, size = 1, lty = 4) + 
        annotate(geom = "text", 
                 label = expression(beta[1]==1~~~beta[0]==2), 
                 x = 1, y = 3.3, parse = TRUE, 
                 angle = 60) +
        annotate(geom = "text", 
                 label = expression(beta[1]==1~~~beta[0]==1), 
                 x = 1, y = 2.3, parse = TRUE, 
                 angle = 60) +
          annotate(geom = "text", 
                 label = expression(beta[1]==1~~~beta[0]==0), 
                 x = 1, y = 1.3, parse = TRUE,
                 angle = 60) +
        theme_minimal() + scale_x_continuous(limits = c(0, 4), expand = c(0,0)) +
        scale_y_continuous(limits = c(0, 4), expand = c(0,0)) +
        theme(axis.title.y = element_blank(),
              axis.text.y = element_blank()) +
         xlab("Dependent variable (x)") + 
        theme(panel.background = element_rect(color = "black"))

c <- ggplot(data = data.frame(x = c(0, 5), y = c(0, 5)), aes(x, y)) + geom_point() +
        geom_abline(slope = 3, intercept = 1, size = 1, lty = 2) + 
        geom_abline(slope = -1, intercept = 3, size = 1, lty = 3) + 
        geom_abline(slope = -4, intercept = 4, size = 1, lty = 4) + 
        annotate(geom = "text", 
                 label = "A", 
                 x = 1.15, y = 0.5, parse = TRUE) +
        annotate(geom = "text", 
                 label = "B", 
                 x = 2, y = 1.5, parse = TRUE) +
          annotate(geom = "text", 
                 label = "C",
                 x = 1.15, y = 3.8, parse = TRUE) +

        theme_minimal() + scale_x_continuous(limits = c(0, 4), expand = c(0,0)) +
        scale_y_continuous(limits = c(0, 4), expand = c(0,0)) +
         theme(axis.title.y = element_blank(),
              axis.text.y = element_blank()) +
        xlab("") + 
        theme(panel.background = element_rect(color = "black"))

plot_grid(a, b, c, ncol = 3)


```

# Estimating the best fit line

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height=4, fig.align="center"}

df <- data.frame(x = c(4,6,8,12,15,23,22),
           Observed = c(5,8,9,9,13,25,34), obs = c("a", "b", "c", "d", "e", "f", "g"))

mod <- lm(Observed ~ x, data = df)

df$Predicted <- predict(mod)

df %>%
        gather(y.val, y, c(Observed,Predicted)) %>%
        ggplot(aes(x, y, group = obs, color = y.val)) + 
         geom_line(color = "green") +
                geom_abline(intercept = coef(mod)[1], slope = coef(mod)[2])+
           geom_point() + scale_color_manual(values = c("blue", "red"), name = "") +
       theme_minimal() + xlab("Independent variable (x)") + ylab("Dependent variable (y)")

```

# Estimating the best fit line

* The best fit line can be estimated by minimizing the vertical distances between **observed** and **predicted** values.
* The distance between bobserved and predicted values are called **residuals**, these can help us *diagnose* the regression.
* The **residuals** are also used to *estimate* the standard errors of the parameters in the model. 

# The standard error of the regression parameter is an estimate of the SD of the sampling distribution 
```{r,  echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height=3.8, fig.width=5, cache=TRUE}

set.seed(2)


N <- 1000000 # number of pairs in simulation

# Create some randomly distributed X values representing some quantrity

X <- runif(N, 100, 5000) / 1000

# generate some y values according to a linear model. Here, they have
# slope = 1.2, intercept = 1 and Gaussian (normal) random noise added with
# mean = 0 and standard deviation = 1. 


Y <- 1.2*X + 1 + rnorm(N,0,1)


## Create a sampling distribution of the intercept and slpe term

n <- 20 # sample size

results <- data.frame(intercept = rep(NA, 10), 
                      slope = rep(NA, 10), 
                      intercept.se = rep(NA, 10), 
                      slope.se = rep(NA, 10))

for(i in 1:10000) {
        
      samp <-   sample(1:N, n, replace = FALSE)
        
      df <- data.frame(y = Y[samp], x = X[samp])  
        
      mod <- lm(y ~ x, data = df)
      
      results[i, 1] <- coef(mod)[1]
      results[i, 2] <- coef(mod)[2]
      results[i, 3] <- coef(summary(mod))[1, 2]
      results[i, 4] <- coef(summary(mod))[2, 2]
     
}


est <- results %>%
        summarise(intercept.sd = sd(intercept),
                  slope.sd = sd(slope), 
                  intercept.se = mean(intercept.se),
                  slope.se = mean(slope.se))


intercepts <- results %>%
        ggplot(aes(intercept)) + geom_density() + theme_minimal() +
        ggtitle("Distribution of sample intercepts (n = 20)") + 
        ylab("Density") + xlab("Intercept") + 
        annotate(geom = "text",x = 2.2, y = 0.6, label = paste("SD = ", round(est[1, 1], 2))) +
        annotate(geom = "text",x = 2.2, y = 0.5, label = paste("Mean SE = ", round(est[1, 3], 2)))

slopes <- results %>%
        ggplot(aes(slope)) + geom_density() + theme_minimal() +
        ggtitle("Distribution of sample slopes (n = 20)") + 
        ylab("Density") + xlab("Slope") +
        annotate(geom = "text",x = 1.6, y = 1.8, label = paste("SD = ", round(est[1, 2], 2))) +
        annotate(geom = "text",x = 1.6, y = 1.5, label = paste("Mean SE = ", round(est[1, 4], 2)))


library(cowplot)

plot_grid(intercepts, slopes, ncol = 1)

```

# Assumptions in linear regression

* There is a linear relationship between $x$ and $y$
* Residuals are normally distributed (with mean = 0)
* Residuals have an equal spread along the the fitted range (homoskedasticity)
* Observations are independent


# Why are assumptions important

* We assume taht errors in our model ($\hat{y}-y$) are well behaved
* The errors are used to calculate standard errors
* If the assumptions are wrong our standard errors are **biased**
* *Biased* standard errors will lead to bad **inference**

# Model diagnostics

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height=4, fig.width=6, cache=TRUE}

set.seed(666)


N <- 50 # number of pairs in simulation

# Create some randomly distributed X values representing some quantrity
x <- runif(N, 100, 5000) / 1000
df <- data.frame(x = x,
                 y = 25*x + 110 + rnorm(N,0,5), 
                 y2 = 25*x + 110 + rnorm(N,0,5)*x)

model <- lm(y ~ x, data = df) 
model2 <- lm(y2 ~ x, data = df)


df$resid <- resid(model)
df$resid2 <- resid(model2)
df$fitted <-fitted(model)
df$fitted2 <- fitted(model2)




a <- df %>%
        ggplot(aes(x, y)) + geom_point() +  geom_smooth(method = "lm")+
        theme_minimal() + xlab("Dependent variable (x)") + 
        ylab("Independent variable (y)")


b <- df %>%
        ggplot(aes(fitted, resid)) + geom_point() +
        theme_minimal() + xlab("Fitted values") + 
        ylab(expression(paste("Raw residuals (",hat(y)-y,")"))) +
        geom_hline(yintercept = 0)

c <- df %>%
        ggplot(aes(x, y2)) + geom_point() +  geom_smooth(method = "lm")+
        theme_minimal() + xlab("Dependent variable (x)") + 
        ylab("Independent variable (y)")


d <- df %>%
        ggplot(aes(fitted2, resid2)) + geom_point() +
        theme_minimal() + xlab("Fitted values") + 
        ylab(expression(paste("Raw residuals (",hat(y)-y,")"))) +
        geom_hline(yintercept = 0)


plot_grid(a, b, c, d, ncol = 2)

DATASET <- df[,c(1,2)]

```

# What can be done with heteroscedasticity?

* Transformation of the data can reduce increased variation with increased values
* The most common transformation is the log
* log-transformed data 

```{r, eval = FALSE}
df$y <- log(df$y)
```

# Log-transfomed data

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height=4, fig.width=6, cache=TRUE}

set.seed(134)


N <- 50 # number of pairs in simulation

# Create some randomly distributed X values representing some quantrity
x <- runif(N, 200, 5000)/1000
df <- data.frame(x = x,
                 y = 25*x + 110 + rnorm(N,0,5), 
                 y2 = 25*x + 110 + runif(N,0,20) * x)

model <- lm(y ~ x, data = df) 
model2 <- lm(y2 ~ x, data = df)
model3 <- lm(log(y2) ~ x, data = df)


df$resid3 <- resid(model3) / sd(resid(model3))
df$resid2 <- resid(model2) / sd(resid(model2))

df$fitted3 <-fitted(model3)
df$fitted2 <- fitted(model2)






c <- df %>%
        ggplot(aes(x, y2)) + geom_point() +  geom_smooth(method = "lm")+
        theme_minimal() + xlab("Dependent variable (x)") + 
        ylab("Independent variable (y)")


d <- df %>%
        ggplot(aes(fitted2, resid2)) + geom_point() +
        theme_minimal() + xlab("Fitted values") + 
        ylab("Standardized residuals") +
        geom_hline(yintercept = 0)

e <- df %>%
        ggplot(aes(x, log(y2))) + geom_point() +  geom_smooth(method = "lm")+
        theme_minimal() + xlab("Dependent variable (x)") + 
        ylab("Independent variable log(y)")


f <- df %>%
        ggplot(aes(fitted3, resid3)) + geom_point() +
        theme_minimal() + xlab("Fitted values") + 
        ylab("Standardized residuals") +
        geom_hline(yintercept = 0)


plot_grid(c, d,e, f, ncol = 2)

DATASET <- df[,c(1,2)]

```

# Interpreting log-transfomed data in a regression

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align='center', fig.height=4, fig.width=6, cache=TRUE}
library(tidyverse)
set.seed(2)
df <- data.frame(a = rnorm(25, 125, 10),
                 b = rnorm(25, 180, 40)) %>%
        
        pivot_longer(names_to = "group", 
                     values_to = "value", 
                     cols = a:b) %>%
        mutate(value = value + runif(50, 1, 10) * value)


m1 <- lm(value~group, data = df)
m2 <- lm(log(value)~group, data = df)

df$resid_m1 <- resid(m1)/sd(resid(m1))
df$resid_m2 <- resid(m2)/sd(resid(m2))
df$fitted_m1 <- fitted(m1)
df$fitted_m2 <- fitted(m2)



c <-  df %>%
        ggplot(aes(group, value, fill = group)) + 
        geom_point(position = position_jitter(width = 0.1), shape = 21, size = 3) + 
        labs(x = "Group", 
             y = "Linear scale value") + theme_minimal() +
        scale_fill_manual(values = c("lightblue", "coral2")) +
        theme(legend.position = "none")


d <-  df %>%
        ggplot(aes(group, log(value), fill = group)) + 
        geom_point(position = position_jitter(width = 0.1), shape = 21, size = 3) + 
        labs(x = "Group", 
             y = "Log scale value") + theme_minimal() +
        scale_fill_manual(values = c("lightblue", "coral2")) +
        theme(legend.position = "none")





a <- df %>%
        ggplot(aes(fitted_m1, resid_m1, fill = group)) + 
        geom_point(position = position_jitter(width = 10), shape = 21, size = 3) + 
        scale_y_continuous(breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), limits = c(-3.5, 3.5)) +
        labs(x = "Fitted values", 
             y = "Standardized residuals", 
             title = "Linear scale") + theme_minimal() +
        scale_fill_manual(values = c("lightblue", "coral2")) +
        theme(legend.position = "none")
b <- df %>%
        ggplot(aes(fitted_m2, resid_m2, fill = group)) + 
        geom_point(position = position_jitter(width = 0.01), shape = 21, size = 3) + 
        scale_y_continuous(breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), limits = c(-3.5, 3.5)) +
        labs(x = "Fitted values", 
             y = "Standardized residuals",
             title = "Log-scale") + theme_minimal() +
        scale_fill_manual(values = c("lightblue", "coral2")) +
        theme(legend.position = "none")


cowplot::plot_grid(c, d, a, b, ncol = 2)

```

# Interpreting log-transfomed data in a regression

```{r, results = "asis", message=FALSE, warning = FALSE, echo = FALSE}
library(broom); library(knitr)

tidy(m1) %>%
        mutate(Model = "Linear scale") %>%
        rbind(tidy(m2) %>%
                      mutate(Model = "Log-transformed")) %>%
        kable(format = "latex", booktabs = TRUE, col.names = c("Paramter", "Estimate", "SE", "t-value", "p-value", "Model"), 
              digits = c(NA, 2, 3, 2, 3)) 
```


$$ log(a) + log(b) = log(a \times b)$$
$$ log(a) - log(b) = log(a/b)$$
$$e^{log(x)} = x$$



# Categorical data can be used as predictor variables

* We can use categorical data as the independent variable
* Categories are (automatically in R) converted to "dummy varaibles"
* If we have two groups (e.g. men and women), in the univariate model, men will be represented by the intercept and women by the slope. 

$$y = \beta_0 + \beta_1 x + \epsilon$$

$$y = MEN + \beta_1 \times WOMEN + \epsilon$$

* If there are more levels, additional dummy variables are added to the model